---
title: "ashers census project"
output: html_document
date: "2026-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Overview

Notes:
-change census api key install = to TRUE if you want to install my API key. or use your own idk if it matters
-the geography csv just provides a list of counties
-right now it's looking at non-combustible energy as a share of total energy production (both columns that just exist in the egrid dataset 😍)
-you can't have the excel file open while the program is looking at it (workbooks live in the `egrid files` folder)


1. Setup

```{r libraries, message=FALSE, warning=FALSE}
library(tidycensus)
library(tidyverse)
library(sf)
library(tigris)
library(readxl)
library(janitor)
library(spdep)
library(ggplot2)

options(tigris_use_cache = TRUE)

# CONFIGURE GEOGRAPHY HERE
# Options: "arc" (Appalachian Regional Commission), 
#          "msa" (Metropolitan Statistical Areas), 
#          "state" (individual states), etc.
#geography_type <- "subregions"
geography_file <- "geography.csv"  # File with county list for selected geography

# CONFIGURE SUBREGION HERE
# Change this to analyze different subregions
#target_subregion <- "Central"

# CONFIGURE ANALYSIS YEAR HERE
# Change this to analyze different years (must be available in both ACS and eGRID)
analysis_year <- 2021

# Baseline year for "change in non-fossil share" analysis (end year is analysis_year)
egrid_change_baseline_year <- 2016

```

2. Load Geography Data
We load selected geography's county list and filter by subregion.
```{r}
geography_df <- read_csv(geography_file) |>
  clean_names() |>
#  filter(subregion == target_subregion) |>
  mutate(fips = str_pad(fips, 5, pad = "0"))
```

3. Income Data (ACS)
We use median household income (B19013_001).
Why two versions?
Aligned (analysis_year): Matches eGRID data → best for correlation
Lagged (analysis_year-2): Income precedes energy → tests robustness
```{r}
census_api_key("0b99bc09077560a4ae65416a235a4f0910a40637", install = FALSE)

# Aligned dataset
income_aligned <- get_acs(
  geography = "county",
  variables = "B19013_001",
  year = analysis_year,
  survey = "acs5"
) |>
  select(GEOID, estimate) |>
  rename(median_income = estimate)

# Lagged dataset (sensitivity test)
income_lagged <- get_acs(
  geography = "county",
  variables = "B19013_001",
  year = analysis_year - 2,
  survey = "acs5"
) |>
  select(GEOID, estimate) |>
  rename(median_income = estimate)

# Restrict to selected geography
income_aligned_app <- income_aligned |>
  inner_join(geography_df, by = c("GEOID" = "fips"))

income_lagged_app <- income_lagged |>
  inner_join(geography_df, by = c("GEOID" = "fips"))
```

4. eGRID Data (Energy Production)
We load plant-level generation for `analysis_year`, aggregate to county, and compute non-combustion + biomass share. Workbooks live in `egrid files/` as `eGRID{year}_data.xlsx`. The plant sheet is either `PLNT{yy}` or `EGRDPLNT{yy}` (`yy` = last two digits of the year).
```{r}
load_egrid_year <- function(year) {
  file_name <- file.path("egrid files", paste0("eGRID", year, "_data.xlsx"))
  sheets <- excel_sheets(file_name)
  yy <- sprintf("%02d", year %% 100)
  sheet_plnt <- paste0("PLNT", yy)
  sheet_egrd <- paste0("EGRDPLNT", yy)
  sheet_name <- if (sheet_plnt %in% sheets) {
    sheet_plnt
  } else if (sheet_egrd %in% sheets) {
    sheet_egrd
  } else {
    stop(
      "No plant sheet ", sheet_plnt, " or ", sheet_egrd, " in ", file_name,
      ". Available: ", paste(sheets, collapse = ", ")
    )
  }

  plants <- read_excel(file_name, sheet = sheet_name, skip = 1) |>
    clean_names() |>
    filter(!is.na(fipsst), !is.na(fipscnty), plngenan > 0) |>
    mutate(
      fips = str_c(str_pad(fipsst, 2, pad = "0"), str_pad(fipscnty, 3, pad = "0")),
      total_gen = plngenan,
      non_fossil_gen = plgenacn + plgenabm,
      year = year
    )

  county_energy_year <- plants |>
    group_by(fips, year) |>
    summarise(
      total_gen = sum(total_gen, na.rm = TRUE),
      non_fossil_gen = sum(non_fossil_gen, na.rm = TRUE)
    ) |>
    mutate(non_fossil_share = non_fossil_gen / total_gen) |>
    filter(total_gen > 0)

  county_energy_year
}

county_energy <- load_egrid_year(analysis_year)
```

5. Build Analysis Datasets
We combine income and energy data.
```{r}
analysis_aligned <- income_aligned_app |>
  inner_join(county_energy, by = c("GEOID" = "fips")) |>
  drop_na()

analysis_lagged <- income_lagged_app |>
  inner_join(county_energy, by = c("GEOID" = "fips")) |>
  drop_na()
```

6. Correlation and Regression
We estimate:
Pearson correlation
Linear regression
```{r}
# run_analysis <- function(df, label) {
#   cat("\n============================\n")
#   cat("RESULTS:", label, "\n")
#   cat("============================\n")
# 
#   cor_res <- cor.test(
#     df$median_income,
#     df$non_fossil_share,
#     method = "pearson"
#   )
#   print(cor_res)
# 
#   model <- lm(non_fossil_share ~ median_income, data = df)
#   print(summary(model))
# }
# 
# run_analysis(analysis_aligned, paste("Aligned (", analysis_year, ")"))
# run_analysis(analysis_lagged, paste("Lagged (", analysis_year-2, ")"))
```

7. Visualization
```{r}
# plot_fn <- function(df, title) {
#   ggplot(df, aes(x = median_income, y = non_fossil_share)) +
#     geom_point(alpha = 0.3) +
#     geom_smooth(method = "lm", se = TRUE) +
#     labs(
#       title = title,
#       x = "Median Household Income",
#       y = "Non-Fossil Energy Share"
#     ) +
#     theme_minimal()
# }
# 
# plot_fn(analysis_aligned, paste("Aligned Income vs Energy (", analysis_year, ")"))
# plot_fn(analysis_lagged, paste("Lagged Income vs Energy (", analysis_year-2, ")"))
```

8. Change in non-fossil share vs median household income
We take eGRID non-combustion share at `egrid_change_baseline_year` and at `analysis_year`, compute the county-level change, and correlate it with median household income from `income_aligned` (same object as above: ACS `B19013_001` for `analysis_year`), restricted to the selected geography via `income_aligned_app`.
```{r}
county_energy_baseline <- load_egrid_year(egrid_change_baseline_year)

energy_share_change <- county_energy_baseline |>
  select(fips, non_fossil_share) |>
  rename(share_baseline = non_fossil_share) |>
  inner_join(
    county_energy |> select(fips, non_fossil_share) |> rename(share_end = non_fossil_share),
    by = "fips"
  ) |>
  mutate(non_fossil_share_change = share_end - share_baseline)

analysis_energy_change <- income_aligned_app |>
  inner_join(energy_share_change, by = c("GEOID" = "fips")) |>
  drop_na(median_income, non_fossil_share_change)
```

```{r}
cat("\n============================\n")
cat(
  "CHANGE ANALYSIS: non-fossil share (",
  egrid_change_baseline_year,
  " → ",
  analysis_year,
  ") vs median household income (ACS ",
  analysis_year,
  ", B19013_001)\n",
  sep = ""
)
cat("============================\n")

cor_change <- cor.test(
  analysis_energy_change$median_income,
  analysis_energy_change$non_fossil_share_change,
  method = "pearson"
)
print(cor_change)

model_change <- lm(non_fossil_share_change ~ median_income, data = analysis_energy_change)
print(summary(model_change))
```

```{r}
ggplot(analysis_energy_change, aes(x = median_income, y = non_fossil_share_change)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = paste0(
      "Change in non-fossil share (",
      egrid_change_baseline_year,
      "–",
      analysis_year,
      ") vs median household income (",
      analysis_year,
      " ACS)"
    ),
    x = "Median household income (B19013_001)",
    y = "Change in non-fossil generation share"
  ) +
  theme_minimal()
```

9. Choropleth: change in non-fossil generation (MWh)

Counties are restricted to `geography_df` and to counties with eGRID plant data in both `egrid_change_baseline_year` and `analysis_year` (same filters as `load_egrid_year`). Fill is end-year minus baseline non-fossil generation (MWh).

```{r map-nf-gen-change-build}
state_fips <- geography_df |>
  mutate(state_fips = str_sub(fips, 1, 2)) |>
  distinct(state_fips) |>
  pull(state_fips)

counties_sf <- counties(state = state_fips, cb = TRUE, year = 2020) |>
  st_transform(crs = 5070)

chg_nf_gen <- geography_df |>
  select(fips) |>
  inner_join(
    county_energy_baseline |>
      select(fips, non_fossil_gen) |>
      rename(non_fossil_gen_baseline = non_fossil_gen),
    by = "fips"
  ) |>
  inner_join(
    county_energy |>
      select(fips, non_fossil_gen) |>
      rename(non_fossil_gen_end = non_fossil_gen),
    by = "fips"
  ) |>
  mutate(non_fossil_gen_change = non_fossil_gen_end - non_fossil_gen_baseline)

counties_map_nf <- counties_sf |>
  inner_join(chg_nf_gen, by = c("GEOID" = "fips"))
```

```{r map-nf-gen-change-plot}
ggplot(counties_map_nf, aes(fill = non_fossil_gen_change)) +
  geom_sf(color = "grey35", linewidth = 0.08) +
  scale_fill_gradient2(
    low = "#b35806",
    mid = "#f7f7f7",
    high = "#542788",
    midpoint = 0,
    name = "Δ Non-fossil\ngeneration (MWh)",
    labels = scales::label_comma()
  ) +
  theme_void() +
  theme(legend.position = "right") +
  labs(
    title = paste0(
      "Change in non-fossil generation (MWh), ",
      egrid_change_baseline_year,
      "–",
      analysis_year
    ),
    subtitle = "Counties in selected geography with eGRID plants in both years",
    caption = "Non-fossil generation from eGRID (plant-level), summed by county."
  )
```
