---
title: "ashers census project"
output: html_document
date: "2026-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Overview

Notes:
-change census api key install = to TRUE if you want to install my API key. or use your own idk if it matters
-the geography csv just provides a list of counties
-right now it's looking at plant annual CO2 emissions (`plco2an`) summed to a county level
-you can't have the excel file open while the program is looking at it (workbooks live in the `egrid files` folder)


1. Setup

```{r libraries, message=FALSE, warning=FALSE}
library(tidycensus)
library(tidyverse)
library(sf)
library(tigris)
library(readxl)
library(janitor)
library(spdep)
library(ggplot2)
library(lme4)
library(emmeans)

options(tigris_use_cache = TRUE)

# CONFIGURE GEOGRAPHY HERE
# Options: "arc" (Appalachian Regional Commission), 
#          "msa" (Metropolitan Statistical Areas), 
#          "state" (individual states), etc.
#geography_type <- "subregions"
geography_file <- "geography.csv"  # File with county list for selected geography

# CONFIGURE SUBREGION HERE
# Change this to analyze different subregions
#target_subregion <- "Central"

# CONFIGURE ANALYSIS YEAR HERE
# Change this to analyze different years (must be available in both ACS and eGRID)
analysis_year <- 2022

# Baseline year for change analysis (end year is analysis_year)
egrid_change_baseline_year <- 2012

# ENERGY VARIABLES (all three run in one knit)
#  - "renewable_mwh": change in renewable generation (MWh) (plgenacn + plgenabm)
#  - "co2_tons":      change in CO2 emissions (tons) (plco2an)
#  - "COAL":          change in generation from coal (MWh) (plgenacl)
energy_vars <- c("renewable_mwh", "co2_tons", "COAL")

# EXCLUDED COUNTIES (outliers for renewable_mwh analysis)
# These 4 counties are removed from all analyses due to extreme renewable energy increases
excluded_counties <- c("47143", "01083", "42079", "47065")

energy_label_for <- function(energy_var) {
  if (energy_var == "renewable_mwh") {
    "Change in renewable energy generation (MWh)"
  } else if (energy_var == "co2_tons") {
    "Change in annual CO2 emissions (tons)"
  } else {
    "Change in generation from coal (MWh)"
  }
}

```

2. Load Geography Data
We load selected geography's county list and filter by subregion.
```{r}
geography_df <- read_csv(geography_file) |>
  clean_names() |>
#  filter(subregion == target_subregion) |>
  mutate(fips = str_pad(fips, 5, pad = "0")) |>
  filter(!fips %in% excluded_counties)

cat("Excluded counties from analysis:", paste(excluded_counties, collapse = ", "), "\n")
cat("Remaining counties in geography_df:", nrow(geography_df), "\n")
```

3. Income Data (ACS)
We use median household income (B19013_001).
Why two versions?
Aligned (analysis_year): Matches eGRID data → best for correlation
Lagged (analysis_year-2): Income precedes energy → tests robustness

ACS pulls are stored in `census_cache` so re-running chunks in the same R session does not repeat API calls (clear `census_cache` or restart R if you change `analysis_year`, `egrid_change_baseline_year`, or other ACS settings).

Metro/micro CBSA boundaries for section 9 are also fetched here.
```{r}
census_api_key("0b99bc09077560a4ae65416a235a4f0910a40637", install = FALSE)

if (!exists("census_cache", envir = .GlobalEnv)) {
  assign("census_cache", new.env(parent = emptyenv()), envir = .GlobalEnv)
}

if (!exists("income_aligned", envir = census_cache, inherits = FALSE)) {
  income_analysis <- get_acs(
    geography = "county",
    variables = "B19013_001",
    year = analysis_year,
    survey = "acs5"
  ) |>
    select(GEOID, estimate) |>
    rename(median_income = estimate)

  income_baseline <- get_acs(
    geography = "county",
    variables = "B19013_001",
    year = egrid_change_baseline_year,
    survey = "acs5"
  ) |>
    select(GEOID, estimate) |>
    rename(median_income_baseline = estimate)

  assign(
    "income_aligned",
    income_analysis |>
      left_join(income_baseline, by = "GEOID") |>
      mutate(median_income_change = median_income - median_income_baseline),
    envir = census_cache
  )
}

if (!exists("income_lagged", envir = census_cache, inherits = FALSE)) {
  assign(
    "income_lagged",
    get_acs(
      geography = "county",
      variables = "B19013_001",
      year = analysis_year - 2,
      survey = "acs5"
    ) |>
      select(GEOID, estimate) |>
      rename(median_income = estimate),
    envir = census_cache
  )
}

if (!exists("cbsa_sf", envir = census_cache, inherits = FALSE)) {
  assign(
    "cbsa_sf",
    get_acs(
      geography = "metropolitan statistical area/micropolitan statistical area",
      variables = "B01001_001",
      year = analysis_year,
      survey = "acs5",
      geometry = TRUE
    ) |>
      st_transform(crs = 5070) |>
      mutate(
        cbsa_type = case_when(
          str_detect(NAME, "Metro Area") ~ "Metropolitan",
          str_detect(NAME, "Micro Area") ~ "Micropolitan",
          TRUE ~ NA_character_
        )
      ) |>
      select(cbsa_type),
    envir = census_cache
  )
}

income_aligned <- get("income_aligned", envir = census_cache)
income_lagged <- get("income_lagged", envir = census_cache)
cbsa_sf <- get("cbsa_sf", envir = census_cache)

# Restrict to selected geography
income_aligned_app <- income_aligned |>
  inner_join(geography_df, by = c("GEOID" = "fips"))

income_lagged_app <- income_lagged |>
  inner_join(geography_df, by = c("GEOID" = "fips"))
```

4. eGRID Data (Energy Change)
We load plant-level eGRID data for the baseline and analysis years, aggregate to county, and compute change for each entry in `energy_vars`. Workbooks live in `egrid files/` as `eGRID{year}_data.xlsx`. The plant sheet is either `PLNT{yy}` or `EGRDPLNT{yy}` (`yy` = last two digits of the year).
```{r}
load_egrid_year <- function(analysis_year, baseline_year, energy_var = c("renewable_mwh", "co2_tons", "COAL")) {
  energy_var <- match.arg(energy_var)

  # Read one year and normalize county FIPS.
  load_plants_for_year <- function(year, energy_var) {
    file_name <- file.path("egrid files", paste0("eGRID", year, "_data.xlsx"))
    sheets <- excel_sheets(file_name)
    yy <- sprintf("%02d", year %% 100)
    sheet_plnt <- paste0("PLNT", yy)
    sheet_egrd <- paste0("EGRDPLNT", yy)
    sheet_name <- if (sheet_plnt %in% sheets) {
      sheet_plnt
    } else if (sheet_egrd %in% sheets) {
      sheet_egrd
    } else {
      stop(
        "No plant sheet ", sheet_plnt, " or ", sheet_egrd, " in ", file_name,
        ". Available: ", paste(sheets, collapse = ", ")
      )
    }

    plants <- read_excel(file_name, sheet = sheet_name, skip = 1) |>
      clean_names() |>
      filter(!is.na(fipsst), !is.na(fipscnty)) |>
      mutate(
        fips = str_c(str_pad(fipsst, 2, pad = "0"), str_pad(fipscnty, 3, pad = "0"))
      )

    if (energy_var == "renewable_mwh") {
      plants <- plants |>
        mutate(energy_value = plgenacn + plgenabm)
    } else if (energy_var == "co2_tons") {
      plants <- plants |>
        mutate(energy_value = plco2an)
    } else if (energy_var == "COAL") {
      plants <- plants |>
        mutate(energy_value = plgenacl)
    }

    plants |>
      mutate(
        is_energy_producing = !is.na(plngenan) & plngenan > 0,
        energy_value = replace_na(energy_value, 0)
      ) |>
      group_by(fips) |>
      summarise(
        energy = sum(energy_value, na.rm = TRUE),
        is_energy_producing = any(is_energy_producing, na.rm = TRUE),
        .groups = "drop"
      )
  }

  county_baseline <- load_plants_for_year(baseline_year, energy_var)
  county_analysis <- load_plants_for_year(analysis_year, energy_var)

  producing_fips <- full_join(
    county_baseline |> select(fips, prod_base = is_energy_producing),
    county_analysis |> select(fips, prod_end = is_energy_producing),
    by = "fips"
  ) |>
    mutate(
      prod_base = replace_na(prod_base, FALSE),
      prod_end = replace_na(prod_end, FALSE),
      include_county = prod_base | prod_end
    ) |>
    filter(include_county) |>
    select(fips)

  county_change <- producing_fips |>
    left_join(
      county_baseline |>
        select(fips, energy_baseline = energy),
      by = "fips"
    ) |>
    left_join(
      county_analysis |>
        select(fips, energy_end = energy),
      by = "fips"
    ) |>
    mutate(
      energy_baseline = replace_na(energy_baseline, 0),
      energy_end = replace_na(energy_end, 0)
    ) |>
    mutate(energy_change = energy_end - energy_baseline)

  county_change
}

county_energy_by_var <- energy_vars |>
  set_names() |>
  map(function(ev) load_egrid_year(analysis_year, egrid_change_baseline_year, ev))

# County-year panel across all available eGRID years in analysis period.
load_egrid_panel <- function(baseline_year, analysis_year, energy_var = c("renewable_mwh", "co2_tons", "COAL")) {
  energy_var <- match.arg(energy_var)

  egrid_files <- list.files(
    path = "egrid files",
    pattern = "^eGRID[0-9]{4}_data\\.xlsx$",
    full.names = FALSE
  )

  available_years <- egrid_files |>
    str_extract("[0-9]{4}") |>
    as.integer() |>
    discard(is.na) |>
    sort()

  years_in_period <- available_years[available_years >= baseline_year & available_years <= analysis_year]

  if (length(years_in_period) == 0) {
    stop(
      "No eGRID files found for years between ", baseline_year, " and ", analysis_year, "."
    )
  }

  map_dfr(years_in_period, function(y) {
    load_egrid_year(y, y, energy_var) |>
      transmute(fips, year = y, e_var = energy_end)
  })
}
```

5. Build Analysis Datasets
For each energy variable, we use the corresponding `county_energy` (with `energy_change`) as the base dataset, then join income data.
```{r}
analysis_county_energy_by_var <- county_energy_by_var |>
  map(function(county_energy) {
    county_energy |>
      inner_join(
        income_aligned_app |>
          select(GEOID, median_income, median_income_change) |>
          rename(median_income_aligned = median_income),
        by = c("fips" = "GEOID")
      ) |>
      left_join(
        income_lagged_app |>
          select(GEOID, median_income) |>
          rename(median_income_lagged = median_income),
        by = c("fips" = "GEOID")
      )
  })

analysis_energy_change_by_var <- analysis_county_energy_by_var |>
  map(~ drop_na(.x, median_income_aligned, energy_change))

```

6. Change in energy variable vs median household income
For each entry in `energy_vars`, we take that eGRID metric at `egrid_change_baseline_year` and at `analysis_year`, compute county-level change inside `load_egrid_year()`, and work with the resulting `energy_change` column.
```{r}
walk(energy_vars, function(ev) {
  analysis_energy_change <- analysis_energy_change_by_var[[ev]]
  cat("\n============================\n")
  cat("CHANGE ANALYSIS [", ev, "]: (", sep = "")
  cat(
    egrid_change_baseline_year,
    " → ",
    analysis_year,
    ") vs median household income (ACS ",
    analysis_year,
    ", B19013_001)\n",
    sep = ""
  )
  cat("============================\n")

  cor_change <- cor.test(
    analysis_energy_change$median_income_aligned,
    analysis_energy_change$energy_change,
    method = "pearson"
  )
  print(cor_change)

  model_change <- lm(energy_change ~ median_income_aligned, data = analysis_energy_change)
  print(summary(model_change))
})
```

```{r}
walk(energy_vars, function(ev) {
  analysis_energy_change <- analysis_energy_change_by_var[[ev]]
  print(
    ggplot(analysis_energy_change, aes(x = median_income_aligned, y = energy_change)) +
      geom_point(alpha = 0.3) +
      geom_smooth(method = "lm", se = TRUE) +
      labs(
        title = paste0(
          "[",
          ev,
          "] Change in energy variable (",
          egrid_change_baseline_year,
          "–",
          analysis_year,
          ") vs median household income (",
          analysis_year,
          " ACS)"
        ),
        x = "Median household income (B19013_001)",
        y = energy_label_for(ev)
      ) +
      theme_minimal()
  )
})
```

6b. Change in energy variable vs change in median household income
For each energy variable, we correlate county-level `energy_change` with county-level `median_income_change` (analysis year minus baseline year income).
```{r}
analysis_income_change_by_var <- energy_vars |>
  set_names() |>
  map(function(ev) {
    analysis_county_energy_by_var[[ev]] |>
      drop_na(median_income_change, energy_change)
  })

walk(energy_vars, function(ev) {
  analysis_income_change <- analysis_income_change_by_var[[ev]]

  cat("\n============================\n")
  cat("INCOME-CHANGE CORRELATION [", ev, "]\n", sep = "")
  cat("============================\n")

  cor_income_change <- cor.test(
    analysis_income_change$median_income_change,
    analysis_income_change$energy_change,
    method = "pearson"
  )
  print(cor_income_change)
})
```

```{r}
walk(energy_vars, function(ev) {
  analysis_income_change <- analysis_income_change_by_var[[ev]]

  print(
    ggplot(analysis_income_change, aes(x = median_income_change, y = energy_change)) +
      geom_point(alpha = 0.3) +
      geom_smooth(method = "lm", se = TRUE) +
      scale_y_continuous(labels = scales::label_comma()) +
      labs(
        title = paste0(
          "[",
          ev,
          "] Change in energy variable (",
          egrid_change_baseline_year,
          "–",
          analysis_year,
          ") vs change in median household income"
        ),
        x = paste0(
          "Change in median household income (ACS ",
          egrid_change_baseline_year,
          " to ",
          analysis_year,
          ")"
        ),
        y = energy_label_for(ev)
      ) +
      theme_minimal() +
      theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "grey30"),
        axis.ticks.x = element_line(color = "grey30")
      )
  )
})
```

7. Choropleth: change in selected energy variable
Fill is `energy_change` from `analysis_energy_change_by_var[[ev]]` for each `ev` in `energy_vars`. A light basemap of all counties in the selected states is drawn underneath for surrounding context.

```{r map-nf-share-change-build}
state_fips <- geography_df |>
  mutate(state_fips = str_sub(fips, 1, 2)) |>
  distinct(state_fips) |>
  pull(state_fips)

counties_sf <- counties(state = state_fips, cb = TRUE, year = 2020) |>
  st_transform(crs = 5070)

states_sf <- states(cb = TRUE, year = 2020) |>
  filter(STATEFP %in% state_fips) |>
  st_transform(crs = 5070)

counties_map_change_by_var <- analysis_energy_change_by_var |>
  map(~ counties_sf |>
    inner_join(
      .x |>
        select(fips, energy_change),
      by = c("GEOID" = "fips")
    ))

geography_outline <- counties_sf |>
  semi_join(geography_df, by = c("GEOID" = "fips")) |>
  st_union() |>
  st_as_sf()
```

```{r map-nf-share-change-plot}
walk(energy_vars, function(ev) {
  counties_map_change <- counties_map_change_by_var[[ev]]
  print(
    ggplot() +
      geom_sf(data = counties_sf, fill = "grey95", color = "grey95", linewidth = 0.05) +
      geom_sf(data = states_sf, fill = NA, color = "grey80", linewidth = 0.25) +
      geom_sf(data = counties_map_change, aes(fill = energy_change), color = "grey35", linewidth = 0.08) +
      geom_sf(data = geography_outline, fill = NA, color = "grey20", linewidth = 0.4) +
      scale_fill_gradient2(
        low = "#0C5000", #swap low and high colors for renewable map
        mid = "#f7f7f7",
        high = "#7F2B0A",
        midpoint = 0,
        name = paste0("Change in \n", ev),
        labels = scales::label_comma()
      ) +
      theme_void() +
      theme(legend.position = "right") +
      labs(
        title = paste0(
          "[",
          ev,
          "] Change in energy variable, ",
          egrid_change_baseline_year,
          "–",
          analysis_year
        ),
        subtitle = "Colored counties are analysis counties; grey counties provide surrounding-area context",
        caption = paste0("Variable: ", ev, "; fill uses energy_change.")
      )
  )
})
```


8. Partisanship vs change in selected energy variable

```{r partisanship-join-and-plot}
# Read partisanship directly from CSV and keep 2022 county-level values.
partisan_join <- read_csv("partisanship.csv", show_col_types = FALSE) |>
  filter(YEAR == 2022) |>
  transmute(
    fips = str_pad(as.character(STCOFIPS10), 5, pad = "0"),
    PARTISAN_INDEX_REP = as.numeric(PARTISAN_INDEX_REP)
  )

if (nrow(partisan_join) == 0) {
  stop("No 2022 rows found in partisanship.csv.")
}

walk(energy_vars, function(ev) {
  analysis_energy_change <- analysis_energy_change_by_var[[ev]]
  analysis_partisanship <- analysis_energy_change |>
    left_join(partisan_join, by = "fips") |>
    drop_na(PARTISAN_INDEX_REP, energy_change)

  cor_partisan <- cor.test(
    analysis_partisanship$PARTISAN_INDEX_REP,
    analysis_partisanship$energy_change,
    method = "pearson"
  )

  cat("\n============================\n")
  cat("PARTISANSHIP CORRELATION [", ev, "]\n", sep = "")
  cat("============================\n")
  print(cor_partisan)

  print(
    ggplot(
      analysis_partisanship,
      aes(x = PARTISAN_INDEX_REP, y = energy_change)
    ) +
      geom_point(alpha = 0.35) +
      geom_smooth(method = "lm", se = TRUE) +
      scale_x_continuous(
        limits = c(0, 1),
        breaks = seq(0, 1, by = 0.2),
        labels = scales::label_number(accuracy = 0.1)
      ) +
      labs(
        title = paste0(
          "[",
          ev,
          "] Partisan index vs change in energy variable (",
          egrid_change_baseline_year,
          "–",
          analysis_year,
          ")"
        ),
        x = "Republican Partisanship (Index 0 to 1)",
        y = energy_label_for(ev)
      ) +
      theme_minimal()
  )
})
```

9. Metro vs micro vs rural (Census CBSA delineations)

```{r metro-micro-mean-sd-plot}
# CBSA polygons from ACS (loaded once in section 3, cached in census_cache)

# Classify counties by whether their interior point lies in a metro/micro CBSA polygon
county_cbsa <- counties_sf |>
  st_point_on_surface() |>
  st_join(cbsa_sf, join = st_within, left = TRUE) |>
  st_drop_geometry() |>
  transmute(
    fips = GEOID,
    cbsa_type = replace_na(cbsa_type, "Rural")
  )
 
walk(energy_vars, function(ev) {
  analysis_energy_change <- analysis_energy_change_by_var[[ev]]
  analysis_cbsa <- analysis_energy_change |>
    left_join(county_cbsa, by = "fips") |>
    mutate(
      county_type = cbsa_type,
      county_type = factor(county_type, levels = c("Metropolitan", "Micropolitan", "Rural"))
    )

  cbsa_summary <- analysis_cbsa |>
    group_by(county_type) |>
    summarise(
      n = n(),
      mean_change = mean(energy_change, na.rm = TRUE),
      sd_change = sd(energy_change, na.rm = TRUE),
      .groups = "drop"
    )

  print(
    ggplot(cbsa_summary, aes(x = county_type, y = mean_change, fill = county_type)) +
      geom_col(width = 0.7, show.legend = FALSE) +
      geom_errorbar(
        aes(ymin = mean_change - sd_change, ymax = mean_change + sd_change),
        width = 0.18,
        linewidth = 0.6
      ) +
      scale_y_continuous(labels = scales::label_comma()) +
      scale_fill_manual(values = c("Metropolitan" = "#ff9900", "Micropolitan" = "#79d8f7", "Rural" = "#6C3400")) +
      labs(
        title = paste0(
          "[",
          ev,
          "] Mean change in energy variable by county CBSA type, ",
          egrid_change_baseline_year,
          "–",
          analysis_year
        ),
        subtitle = "",
        x = "Metro Proximity",
        y = energy_label_for(ev),
        caption = "Error bars show +/- 1 standard deviation"
      ) +
      theme_minimal()
  )
})
```

10. Linear model on asinh-transformed `energy_change`

```{r glmm-energy-change-asinh}
# Ensure partisanship join exists even if section 8 has not run yet.
if (!exists("partisan_join")) {
  partisan_join <- read_csv("partisanship.csv", show_col_types = FALSE) |>
    filter(YEAR == 2022) |>
    transmute(
      fips = str_pad(as.character(STCOFIPS10), 5, pad = "0"),
      partisan_index_rep = as.numeric(PARTISAN_INDEX_REP)
    )
}

partisan_join_glmm <- partisan_join |>
  transmute(
    fips,
    partisan_index_rep = if ("partisan_index_rep" %in% names(partisan_join)) {
      as.numeric(partisan_index_rep)
    } else {
      as.numeric(PARTISAN_INDEX_REP)
    }
  )

# Load hasplan data
hasplan_df <- read_csv("hasplan.csv", show_col_types = FALSE) |>
  clean_names() |>
  transmute(
    fips = str_pad(fips, 5, pad = "0"),
    hasplan_score = factor(score)  # 3 possible states as factor
  )

walk(energy_vars, function(ev) {
  model_df <- analysis_county_energy_by_var[[ev]] |>
    left_join(
      geography_df |>
        select(fips, state),
      by = "fips"
    ) |>
    left_join(
      county_cbsa |>
        transmute(
          fips,
          cbsa_factor = case_when(
            cbsa_type == "Metropolitan" ~ "Metropolitan",
            cbsa_type == "Micropolitan" ~ "Micropolitan",
            TRUE ~ "Rural"
          )
        ),
      by = "fips"
    ) |>
    left_join(partisan_join_glmm, by = "fips") |>
    left_join(hasplan_df, by = "fips") |>
    mutate(
      asinh_energy_change = asinh(energy_change),
      state = factor(state),
      cbsa_factor = factor(cbsa_factor, levels = c("Metropolitan", "Micropolitan", "Rural"))
    ) |>
    drop_na(
      asinh_energy_change,
      median_income_aligned,
      median_income_change,
      partisan_index_rep,
      state,
      cbsa_factor,
      hasplan_score
    )
  
  # Check for counties missing from hasplan dataset
  all_counties <- analysis_county_energy_by_var[[ev]] |> pull(fips) |> unique()
  hasplan_counties <- hasplan_df |> pull(fips) |> unique()
  missing_counties <- setdiff(all_counties, hasplan_counties)
  
  if (length(missing_counties) > 0) {
    cat("\n⚠️ WARNING [", ev, "]: ", length(missing_counties), 
        " counties present in other variables are MISSING from hasplan dataset:\n", sep = "")
    cat(paste(head(missing_counties, 20), collapse = ", "))
    if (length(missing_counties) > 20) {
      cat("... and", length(missing_counties) - 20, "more")
    }
    cat("\n")
  }

  model_df <- model_df |>
    mutate(
      z_median_income_aligned = as.numeric(scale(median_income_aligned)),
      z_median_income_change = as.numeric(scale(median_income_change)),
      z_partisan_index_rep = as.numeric(scale(partisan_index_rep))
    ) |>
    drop_na(
      z_median_income_aligned,
      z_median_income_change,
      z_partisan_index_rep
    )

  # Linear model with all fixed effects (no random effects)
  lm_formula <- asinh_energy_change ~
    z_median_income_aligned +
    z_median_income_change +
    z_partisan_index_rep +
    state +
    cbsa_factor +
    hasplan_score

  lm_fit <- lm(lm_formula, data = model_df)

  partisan_mean <- mean(model_df$partisan_index_rep, na.rm = TRUE)
  partisan_sd <- sd(model_df$partisan_index_rep, na.rm = TRUE)

  partisanship_grid <- c(0, 0.25, 0.5, 0.75, 1)
  partisanship_at <- (partisanship_grid - partisan_mean) / partisan_sd

  cat("\n============================\n")
  cat("LINEAR MODEL [", ev, "]: asinh(energy_change)\n", sep = "")
  cat("============================\n")
  cat(
    "Model: asinh(energy_change) ~ z_median_income_aligned + z_median_income_change + z_partisan_index_rep + state + cbsa_factor + hasplan_score\n"
  )
  cat("Rows used:", nrow(model_df), "\n")
  print(summary(lm_fit))

  cat("\n============================\n")
  cat("EMMEANS (BACK-TRANSFORMED): cbsa_factor [", ev, "]\n", sep = "")
  cat("============================\n")
  emm_cbsa <- emmeans(lm_fit, ~ cbsa_factor)
  emm_cbsa_bt <- summary(emm_cbsa) |>
    as_tibble() |>
    mutate(
      energy_change_est = sinh(emmean),
      energy_change_lower = sinh(lower.CL),
      energy_change_upper = sinh(upper.CL)
    )
  print(emm_cbsa_bt)

  cat("\n============================\n")
  cat("EMMEANS (BACK-TRANSFORMED): partisan_index_rep grid [", ev, "]\n", sep = "")
  cat("============================\n")
  emm_partisan <- emmeans(
    lm_fit,
    ~ z_partisan_index_rep,
    at = list(z_partisan_index_rep = partisanship_at)
  )
  emm_partisan_bt <- summary(emm_partisan) |>
    as_tibble() |>
    mutate(
      partisan_index_rep = partisanship_grid,
      energy_change_est = sinh(emmean),
      energy_change_lower = sinh(lower.CL),
      energy_change_upper = sinh(upper.CL)
    ) |>
    select(
      partisan_index_rep,
      energy_change_est,
      energy_change_lower,
      energy_change_upper,
      everything()
    )
  print(emm_partisan_bt)

  emmeans_plot_theme <- theme_minimal(base_size = 14) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.35),
      panel.grid.minor = element_blank(),
      axis.title = element_text(size = 16, face = "plain"),
      axis.text = element_text(size = 13),
      axis.line.x = element_line(color = "grey35", linewidth = 0.4),
      axis.line.y = element_line(color = "grey35", linewidth = 0.4),
      axis.ticks = element_line(color = "grey28", linewidth = 0.4)
    )

  if (ev == "renewable_mwh") {
    grad_low <- "#2166AC"
    grad_high <- "#B2182B"

    partisan_line_seg <- emm_partisan_bt |>
      mutate(
        xend = lead(partisan_index_rep),
        yend = lead(energy_change_est)
      ) |>
      filter(!is.na(xend)) |>
      mutate(x_mid = (partisan_index_rep + xend) / 2)

    seg_ids <- seq_len(nrow(emm_partisan_bt) - 1)
    partisan_ribbon_poly <- map_dfr(seg_ids, function(i) {
      tibble(
        x = c(
          emm_partisan_bt$partisan_index_rep[i],
          emm_partisan_bt$partisan_index_rep[i + 1],
          emm_partisan_bt$partisan_index_rep[i + 1],
          emm_partisan_bt$partisan_index_rep[i]
        ),
        y = c(
          emm_partisan_bt$energy_change_lower[i],
          emm_partisan_bt$energy_change_lower[i + 1],
          emm_partisan_bt$energy_change_upper[i + 1],
          emm_partisan_bt$energy_change_upper[i]
        ),
        seg_id = i,
        x_mid = mean(c(
          emm_partisan_bt$partisan_index_rep[i],
          emm_partisan_bt$partisan_index_rep[i + 1]
        ))
      )
    })

    print(
      ggplot() +
        geom_polygon(
          data = partisan_ribbon_poly,
          aes(x = x, y = y, group = seg_id, fill = x_mid),
          alpha = 0.45,
          color = NA
        ) +
        scale_fill_gradient(low = grad_low, high = grad_high, limits = c(0, 1), guide = "none") +
        geom_hline(yintercept = 0, color = "grey40", linewidth = 0.45) +
        geom_segment(
          data = partisan_line_seg,
          aes(
            x = partisan_index_rep,
            xend = xend,
            y = energy_change_est,
            yend = yend,
            color = x_mid
          ),
          linewidth = 1,
          lineend = "round"
        ) +
        geom_point(
          data = emm_partisan_bt,
          aes(x = partisan_index_rep, y = energy_change_est, color = partisan_index_rep),
          size = 2.8
        ) +
        scale_color_gradient(low = grad_low, high = grad_high, limits = c(0, 1), guide = "none") +
        scale_x_continuous(
          limits = c(0, 1),
          breaks = seq(0, 1, by = 0.2),
          labels = scales::label_number(accuracy = 0.1),
          expand = expansion(mult = c(0.02, 0.02))
        ) +
        scale_y_continuous(labels = scales::label_comma()) +
        labs(
          x = "Partisan index (0 to 1)",
          y = "Renewable change (MWh)"
        ) +
        emmeans_plot_theme
    )
  }

  if (ev == "COAL") {
    print(
      ggplot(
        emm_cbsa_bt,
        aes(x = cbsa_factor, y = energy_change_est, fill = cbsa_factor)
      ) +
        geom_hline(yintercept = 0, color = "grey40", linewidth = 0.45) +
        geom_col(width = 0.7, show.legend = FALSE) +
        geom_errorbar(
          aes(ymin = energy_change_lower, ymax = energy_change_upper),
          width = 0.18,
          linewidth = 0.65
        ) +
        scale_y_continuous(labels = scales::label_comma()) +
        scale_fill_manual(
          values = c(
            "Metropolitan" = "#ff9900",
            "Micropolitan" = "#79d8f7",
            "Rural/Neither" = "#6C3400"
          )
        ) +
        labs(
          x = "Metro proximity",
          y = "Coal change (MWh)"
        ) +
        emmeans_plot_theme
    )
  }
  
  # ANOVA on the model
  cat("\n============================\n")
  cat("ANOVA on Linear Model\n")
  cat("============================\n")
  print(anova(lm_fit))
  
  # Drop1 test
  cat("\n============================\n")
  cat("Drop1 Test (LRT)\n")
  cat("============================\n")
  print(drop1(lm_fit, test = "Chisq"))

  # Post-hoc Tukey pairwise tests for cbsa_factor
  cat("\n============================\n")
  cat("TUKEY POST-HOC: cbsa_factor [", ev, "]\n", sep = "")
  cat("============================\n")
  
  aov_fit <- aov(lm_formula, data = model_df)
  tukey_cbsa <- TukeyHSD(aov_fit, which = "cbsa_factor")
  print(tukey_cbsa)
  
  # Convert to tibble and show significant results
  tukey_cbsa_tbl <- as.data.frame(tukey_cbsa$cbsa_factor) |>
    rownames_to_column("comparison") |>
    as_tibble() |>
    rename(
      diff = diff,
      conf_low = lwr,
      conf_high = upr,
      p_adj = `p adj`
    ) |>
    arrange(p_adj)
  
  sig_tukey_cbsa <- tukey_cbsa_tbl |>
    filter(p_adj < 0.05)
  
  cat("\nSignificant cbsa_factor pair differences (Tukey-adjusted p < 0.05):\n")
  if (nrow(sig_tukey_cbsa) == 0) {
    cat("None at p < 0.05.\n")
  } else {
    print(sig_tukey_cbsa)
  }
  
  # F-test specifically for cbsa_factor
  cat("\n============================\n")
  cat("F-TEST: cbsa_factor [", ev, "]\n", sep = "")
  cat("============================\n")
  
  # Reduced model without cbsa_factor
  lm_formula_reduced <- asinh_energy_change ~
    z_median_income_aligned +
    z_median_income_change +
    z_partisan_index_rep +
    state +
    hasplan_score
  
  lm_fit_reduced <- lm(lm_formula_reduced, data = model_df)
  
  # Likelihood ratio test
  f_test <- anova(lm_fit_reduced, lm_fit)
  cat("Model comparison (reduced vs full with cbsa_factor):\n")
  print(f_test)
  
  cat("\n---\n")
  
})



```

10a. Post-hoc tests summary

```{r posthoc-summary}
# cat("Post-hoc tests complete for all energy variables.\n")
# cat("See chunk 10 output for Tukey HSD pairwise comparisons and F-tests for cbsa_factor.\n")
```

10b. Distribution of asinh_energy_change and residuals

```{r asinh-energy-dist-residuals}
walk(energy_vars, function(ev) {
  # Get model_df from analysis_county_energy_by_var (recreate minimal needed data)
  model_df <- analysis_county_energy_by_var[[ev]] |>
    left_join(
      geography_df |> select(fips, state),
      by = "fips"
    ) |>
    left_join(
      county_cbsa |>
        transmute(
          fips,
          cbsa_factor = case_when(
            cbsa_type == "Metropolitan" ~ "Metropolitan",
            cbsa_type == "Micropolitan" ~ "Micropolitan",
            TRUE ~ "Rural"
          )
        ),
      by = "fips"
    ) |>
    left_join(
      partisan_join |> transmute(fips, partisan_index_rep = as.numeric(PARTISAN_INDEX_REP)),
      by = "fips"
    ) |>
    left_join(hasplan_df, by = "fips") |>
    mutate(
      asinh_energy_change = asinh(energy_change),
      state = factor(state),
      cbsa_factor = factor(cbsa_factor, levels = c("Metropolitan", "Micropolitan", "Rural"))
    ) |>
    drop_na(asinh_energy_change, median_income_aligned, median_income_change, 
            partisan_index_rep, state, cbsa_factor, hasplan_score)
  
  model_df <- model_df |>
    mutate(
      z_median_income_aligned = as.numeric(scale(median_income_aligned)),
      z_median_income_change = as.numeric(scale(median_income_change)),
      z_partisan_index_rep = as.numeric(scale(partisan_index_rep))
    ) |>
    drop_na(z_median_income_aligned, z_median_income_change, z_partisan_index_rep)
  
  # Fit the same model as chunk 10
  lm_formula <- asinh_energy_change ~
    z_median_income_aligned +
    z_median_income_change +
    z_partisan_index_rep +
    state +
    cbsa_factor +
    hasplan_score
  
  lm_fit <- lm(lm_formula, data = model_df)
  
  # Calculate residuals
  model_df$residuals <- residuals(lm_fit)
  model_df$fitted <- fitted(lm_fit)
  
  # Observed asinh_energy_change stats
  mu_hat <- mean(model_df$asinh_energy_change, na.rm = TRUE)
  sigma_hat <- sd(model_df$asinh_energy_change, na.rm = TRUE)
  
  cat("\n============================\n")
  cat("ASINH_ENERGY_CHANGE DISTRIBUTION [", ev, "]\n", sep = "")
  cat("============================\n")
  cat("n =", nrow(model_df), "\n")
  cat("Observed mean =", round(mu_hat, 4), "\n")
  cat("Observed sd =", round(sigma_hat, 4), "\n")
  
  # Shapiro-Wilk test (if n <= 5000)
  if (nrow(model_df) >= 3 && nrow(model_df) <= 5000) {
    shapiro_obs <- shapiro.test(model_df$asinh_energy_change)
    cat("Shapiro-Wilk p-value (observed):", signif(shapiro_obs$p.value, 4), "\n")
  } else {
    cat("Shapiro-Wilk skipped (requires 3 <= n <= 5000)\n")
  }
  
  # Histogram: observed vs theoretical Gaussian
  print(
    ggplot(model_df, aes(x = asinh_energy_change)) +
      geom_histogram(
        aes(y = after_stat(density)),
        bins = 30,
        fill = "#6AAED6",
        color = "white",
        alpha = 0.85
      ) +
      stat_function(
        fun = dnorm,
        args = list(mean = mu_hat, sd = sigma_hat),
        color = "#B2182B",
        linewidth = 1.2
      ) +
      labs(
        title = paste0("[", ev, "] asinh_energy_change vs Theoretical Gaussian"),
        subtitle = paste0("Blue bars: observed distribution; red curve: N(", round(mu_hat, 2), ", ", round(sigma_hat, 2), ")"),
        x = "asinh(energy_change)",
        y = "Density"
      ) +
      theme_minimal()
  )
  
  # Q-Q plot for observed
  print(
    ggplot(model_df, aes(sample = asinh_energy_change)) +
      stat_qq(alpha = 0.5, color = "#333333") +
      stat_qq_line(color = "#E6550D", linewidth = 0.8) +
      labs(
        title = paste0("[", ev, "] Q-Q plot: asinh_energy_change vs Gaussian"),
        x = "Theoretical quantiles",
        y = "Observed quantiles"
      ) +
      theme_minimal()
  )
  
  # Residuals analysis
  cat("\n============================\n")
  cat("RESIDUALS ANALYSIS [", ev, "]\n", sep = "")
  cat("============================\n")
  cat("Residuals mean =", round(mean(model_df$residuals), 6), "\n")
  cat("Residuals sd =", round(sd(model_df$residuals), 4), "\n")
  
  # Shapiro-Wilk on residuals
  if (nrow(model_df) >= 3 && nrow(model_df) <= 5000) {
    shapiro_resid <- shapiro.test(model_df$residuals)
    cat("Shapiro-Wilk p-value (residuals):", signif(shapiro_resid$p.value, 4), "\n")
  }
  
  # Histogram of residuals
  print(
    ggplot(model_df, aes(x = residuals)) +
      geom_histogram(
        aes(y = after_stat(density)),
        bins = 30,
        fill = "#7FC97F",
        color = "white",
        alpha = 0.85
      ) +
      stat_function(
        fun = dnorm,
        args = list(mean = 0, sd = sd(model_df$residuals)),
        color = "#B2182B",
        linewidth = 1.2
      ) +
      labs(
        title = paste0("[", ev, "] Residuals Distribution"),
        subtitle = paste0("Mean = 0, SD = ", round(sd(model_df$residuals), 2)),
        x = "Residuals",
        y = "Density"
      ) +
      theme_minimal()
  )
  
  # Q-Q plot for residuals
  print(
    ggplot(model_df, aes(sample = residuals)) +
      stat_qq(alpha = 0.5, color = "#333333") +
      stat_qq_line(color = "#E6550D", linewidth = 0.8) +
      labs(
        title = paste0("[", ev, "] Q-Q plot: Residuals vs Gaussian"),
        x = "Theoretical quantiles",
        y = "Observed residuals"
      ) +
      theme_minimal()
  )
  
  # Fitted vs Residuals plot
  print(
    ggplot(model_df, aes(x = fitted, y = residuals)) +
      geom_point(alpha = 0.5) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
      geom_smooth(method = "loess", se = FALSE, color = "blue") +
      labs(
        title = paste0("[", ev, "] Fitted vs Residuals"),
        x = "Fitted values (asinh scale)",
        y = "Residuals"
      ) +
      theme_minimal()
  )
})
```

11. Distribution check: `e_var` and fitted Gaussian model

```{r evar-gaussian-check}
# walk(energy_vars, function(ev) {
#   egrid_panel <- load_egrid_panel(
#     baseline_year = egrid_change_baseline_year,
#     analysis_year = analysis_year,
#     energy_var = ev
#   )
# 
#   dist_df <- egrid_panel |>
#     drop_na(e_var)
# 
#   mu_hat <- mean(dist_df$e_var, na.rm = TRUE)
#   sigma_hat <- sd(dist_df$e_var, na.rm = TRUE)
# 
#   cat("\n============================\n")
#   cat("GAUSSIAN CHECK [", ev, "]\n", sep = "")
#   cat("============================\n")
#   cat("n =", nrow(dist_df), "\n")
#   cat("mean (mu) =", round(mu_hat, 4), "\n")
#   cat("sd (sigma) =", round(sigma_hat, 4), "\n")
# 
#   if (nrow(dist_df) >= 3 && nrow(dist_df) <= 5000) {
#     shapiro_out <- shapiro.test(dist_df$e_var)
#     cat("Shapiro-Wilk p-value =", signif(shapiro_out$p.value, 4), "\n")
#   } else {
#     cat("Shapiro-Wilk skipped (requires 3 <= n <= 5000).\n")
#   }
# 
#   print(
#     ggplot(dist_df, aes(x = e_var)) +
#       geom_histogram(
#         aes(y = after_stat(count)),
#         bins = 35,
#         fill = "#6AAED6",
#         color = "white",
#         alpha = 0.85
#       ) +
#       stat_function(
#         fun = function(x) dnorm(x, mean = mu_hat, sd = sigma_hat) * nrow(dist_df) * (diff(range(dist_df$e_var, na.rm = TRUE)) / 35),
#         color = "#B2182B",
#         linewidth = 1
#       ) +
#       labs(
#         title = paste0("[", ev, "] Distribution of e_var with fitted Gaussian"),
#         subtitle = "Blue bars: observed distribution; red curve: Normal(mu, sigma) fit",
#         x = energy_label_for(ev),
#         y = "Count"
#       ) +
#       theme_minimal()
#   )
# 
#   print(
#     ggplot(dist_df, aes(sample = e_var)) +
#       stat_qq(alpha = 0.35, color = "#333333") +
#       stat_qq_line(color = "#E6550D", linewidth = 0.8) +
#       labs(
#         title = paste0("[", ev, "] Q-Q plot vs Gaussian"),
#         subtitle = "Points close to the line indicate data are approximately Gaussian",
#         x = "Theoretical Gaussian quantiles",
#         y = "Observed quantiles"
#       ) +
#       theme_minimal()
#   )
# })
```

Interpretation guide:
- Histogram + red Gaussian curve: close overlap through center and tails supports a Gaussian model.
- Q-Q plot: points near the line (especially in both tails) support Gaussian assumptions.
- Shapiro-Wilk p-value: a large p-value (commonly > 0.05) means you do not have strong evidence against Gaussianity; a very small p-value suggests non-Gaussian behavior.
- For ANOVA specifically, the normality assumption is mainly about model residuals, so this is an initial data-level check.

12. ANOVA: annual energy variable by year/state/subregion/CBSA

```{r anova-energy-panel}
# anova_df_by_var <- energy_vars |>
#   set_names() |>
#   map(function(ev) {
#     egrid_panel <- load_egrid_panel(
#       baseline_year = egrid_change_baseline_year,
#       analysis_year = analysis_year,
#       energy_var = ev
#     )
# 
#     egrid_panel |>
#       inner_join(
#         geography_df |> select(fips, state, subregion),
#         by = "fips"
#       ) |>
#       left_join(
#         county_cbsa |>
#           transmute(
#             fips,
#             cbsa_factor = case_when(
#               cbsa_type == "Metropolitan" ~ "Metropolitan",
#               cbsa_type == "Micropolitan" ~ "Micropolitan",
#               TRUE ~ "Rural"
#             )
#           ),
#         by = "fips"
#       ) |>
#       mutate(
#         year = factor(year),
#         state = factor(state),
#         subregion = factor(subregion),
#         cbsa_factor = factor(cbsa_factor, levels = c("Metropolitan", "Micropolitan", "Rural"))
#       ) |>
#       drop_na(e_var, year, state, subregion, cbsa_factor)
#   })
# 
# walk(energy_vars, function(ev) {
#   anova_df <- anova_df_by_var[[ev]]
#   anova_model <- lm(e_var ~ year + state + subregion + cbsa_factor, data = anova_df)
#   anova_results <- anova(anova_model)
# 
#   cat("\n============================\n")
#   cat("ANOVA [", ev, "]: e_var ~ year + state + subregion + cbsa_factor\n", sep = "")
#   cat("============================\n")
#   print(anova_results)
# })
```

```{r tukey-posthoc-state}
# walk(energy_vars, function(ev) {
#   anova_df <- anova_df_by_var[[ev]]
# 
#   anova_model_aov <- aov(e_var ~ year + state + subregion + cbsa_factor, data = anova_df)
#   tukey_state <- TukeyHSD(anova_model_aov, which = "state")
# 
#   cat("\n============================\n")
#   cat("TUKEY POST HOC: state [", ev, "]\n", sep = "")
#   cat("============================\n")
#   print(tukey_state)
# 
#   tukey_state_tbl <- as.data.frame(tukey_state$state) |>
#     rownames_to_column("comparison") |>
#     as_tibble() |>
#     rename(
#       diff = diff,
#       conf_low = lwr,
#       conf_high = upr,
#       p_adj = `p adj`
#     ) |>
#     arrange(p_adj)
# 
#   sig_tukey_state <- tukey_state_tbl |>
#     filter(p_adj < 0.05)
# 
#   cat("\nSignificant state pair differences (Tukey-adjusted p < 0.05):\n")
#   if (nrow(sig_tukey_state) == 0) {
#     cat("None at p < 0.05.\n")
#   } else {
#     print(sig_tukey_state)
#   }
# 
#   plot(tukey_state, las = 1)
#   title(main = paste0("Tukey HSD (state): ", ev))
# })
```

```{r tukey-posthoc-cbsa}
# walk(energy_vars, function(ev) {
#   anova_df <- anova_df_by_var[[ev]]
# 
#   anova_model_aov <- aov(e_var ~ year + state + subregion + cbsa_factor, data = anova_df)
#   tukey_cbsa <- TukeyHSD(anova_model_aov, which = "cbsa_factor")
# 
#   cat("\n============================\n")
#   cat("TUKEY POST HOC: cbsa_factor [", ev, "]\n", sep = "")
#   cat("============================\n")
#   print(tukey_cbsa)
# 
#   tukey_cbsa_tbl <- as.data.frame(tukey_cbsa$cbsa_factor) |>
#     rownames_to_column("comparison") |>
#     as_tibble() |>
#     rename(
#       diff = diff,
#       conf_low = lwr,
#       conf_high = upr,
#       p_adj = `p adj`
#     ) |>
#     arrange(p_adj)
# 
#   sig_tukey_cbsa <- tukey_cbsa_tbl |>
#     filter(p_adj < 0.05)
# 
#   cat("\nSignificant cbsa_factor pair differences (Tukey-adjusted p < 0.05):\n")
#   if (nrow(sig_tukey_cbsa) == 0) {
#     cat("None at p < 0.05.\n")
#   } else {
#     print(sig_tukey_cbsa)
#   }
# 
#   plot(tukey_cbsa, las = 1)
#   title(main = paste0("Tukey HSD (cbsa_factor): ", ev))
# })
```

11. Panel model: county and year fixed effects (clustered SE)

```{r panel-fe-model}
# panel_df <- anova_df_by_var[[energy_vars[1]]] |>
#   mutate(
#     county_id = factor(fips),
#     year = factor(year)
#   )
# 
# fe_model <- lm(e_var ~ year + county_id, data = panel_df)
# 
# if (requireNamespace("sandwich", quietly = TRUE) && requireNamespace("lmtest", quietly = TRUE)) {
#   vcov_cluster_county <- sandwich::vcovCL(fe_model, cluster = ~ fips)
#   fe_results_clustered <- lmtest::coeftest(fe_model, vcov. = vcov_cluster_county)
# 
#   cat("\n============================\n")
#   cat("PANEL FE MODEL: e_var ~ year + county fixed effects (SE clustered by county)\n")
#   cat("============================\n")
#   print(fe_results_clustered)
# } else {
#   cat("\n============================\n")
#   cat("PANEL FE MODEL: e_var ~ year + county fixed effects\n")
#   cat("============================\n")
#   cat("Packages 'sandwich' and/or 'lmtest' are not installed; showing default lm() summary.\n")
#   print(summary(fe_model))
# }
# 
# cat("\nNote: With county fixed effects, time-invariant county attributes (state/subregion/CBSA) are absorbed and not separately identified.\n")
```
