---
title: "ashers census project"
output: html_document
date: "2026-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Overview

Notes:
-change census api key install = to TRUE if you want to install my API key. or use your own idk if it matters
-the geography csv just provides a list of counties
-right now it's looking at plant annual CO2 emissions (`plco2an`) summed to a county level
-you can't have the excel file open while the program is looking at it (workbooks live in the `egrid files` folder)


1. Setup

```{r libraries, message=FALSE, warning=FALSE}
library(tidycensus)
library(tidyverse)
library(sf)
library(tigris)
library(readxl)
library(janitor)
library(spdep)
library(ggplot2)

options(tigris_use_cache = TRUE)

# CONFIGURE GEOGRAPHY HERE
# Options: "arc" (Appalachian Regional Commission), 
#          "msa" (Metropolitan Statistical Areas), 
#          "state" (individual states), etc.
#geography_type <- "subregions"
geography_file <- "geography.csv"  # File with county list for selected geography

# CONFIGURE SUBREGION HERE
# Change this to analyze different subregions
#target_subregion <- "Central"

# CONFIGURE ANALYSIS YEAR HERE
# Change this to analyze different years (must be available in both ACS and eGRID)
analysis_year <- 2021

# Baseline year for change analysis (end year is analysis_year)
egrid_change_baseline_year <- 2016

# CONFIGURE ENERGY VARIABLE HERE
# Options:
#  - "renewable_mwh": change in renewable generation (MWh) (plgenacn + plgenabm)
#  - "co2_tons":      change in CO2 emissions (tons) (plco2an)
energy_var <- "renewable_mwh"

energy_label <- if (energy_var == "renewable_mwh") {
  "Change in renewable generation (MWh)"
} else {
  "Change in annual CO2 emissions (tons)"
}

```

2. Load Geography Data
We load selected geography's county list and filter by subregion.
```{r}
geography_df <- read_csv(geography_file) |>
  clean_names() |>
#  filter(subregion == target_subregion) |>
  mutate(fips = str_pad(fips, 5, pad = "0"))
```

3. Income Data (ACS)
We use median household income (B19013_001).
Why two versions?
Aligned (analysis_year): Matches eGRID data → best for correlation
Lagged (analysis_year-2): Income precedes energy → tests robustness
```{r}
census_api_key("0b99bc09077560a4ae65416a235a4f0910a40637", install = FALSE)

# Aligned dataset
income_aligned <- get_acs(
  geography = "county",
  variables = "B19013_001",
  year = analysis_year,
  survey = "acs5"
) |>
  select(GEOID, estimate) |>
  rename(median_income = estimate)

# Lagged dataset (sensitivity test)
income_lagged <- get_acs(
  geography = "county",
  variables = "B19013_001",
  year = analysis_year - 2,
  survey = "acs5"
) |>
  select(GEOID, estimate) |>
  rename(median_income = estimate)

# Restrict to selected geography
income_aligned_app <- income_aligned |>
  inner_join(geography_df, by = c("GEOID" = "fips"))

income_lagged_app <- income_lagged |>
  inner_join(geography_df, by = c("GEOID" = "fips"))
```

4. eGRID Data (Energy Change)
We load plant-level eGRID data for the baseline and analysis years, aggregate to county, and compute change in the selected energy variable (`energy_var`). Workbooks live in `egrid files/` as `eGRID{year}_data.xlsx`. The plant sheet is either `PLNT{yy}` or `EGRDPLNT{yy}` (`yy` = last two digits of the year).
```{r}
load_egrid_year <- function(analysis_year, baseline_year, energy_var = c("renewable_mwh", "co2_tons")) {
  energy_var <- match.arg(energy_var)

  load_plants_for_year <- function(year, energy_var) {
    file_name <- file.path("egrid files", paste0("eGRID", year, "_data.xlsx"))
    sheets <- excel_sheets(file_name)
    yy <- sprintf("%02d", year %% 100)
    sheet_plnt <- paste0("PLNT", yy)
    sheet_egrd <- paste0("EGRDPLNT", yy)
    sheet_name <- if (sheet_plnt %in% sheets) {
      sheet_plnt
    } else if (sheet_egrd %in% sheets) {
      sheet_egrd
    } else {
      stop(
        "No plant sheet ", sheet_plnt, " or ", sheet_egrd, " in ", file_name,
        ". Available: ", paste(sheets, collapse = ", ")
      )
    }

    plants <- read_excel(file_name, sheet = sheet_name, skip = 1) |>
      clean_names() |>
      filter(!is.na(fipsst), !is.na(fipscnty)) |>
      mutate(
        fips = str_c(str_pad(fipsst, 2, pad = "0"), str_pad(fipscnty, 3, pad = "0"))
      )

    if (energy_var == "renewable_mwh") {
      plants <- plants |>
        mutate(energy_value = plgenacn + plgenabm)
    } else if (energy_var == "co2_tons") {
      plants <- plants |>
        mutate(energy_value = plco2an)
    }

    plants |>
      filter(!is.na(energy_value)) |>
      group_by(fips) |>
      summarise(energy = sum(energy_value, na.rm = TRUE), .groups = "drop")
  }

  county_baseline <- load_plants_for_year(baseline_year, energy_var) |>
    rename(energy_baseline = energy)

  county_analysis <- load_plants_for_year(analysis_year, energy_var) |>
    rename(energy_end = energy)

  county_change <- county_baseline |>
    inner_join(county_analysis, by = "fips") |>
    mutate(energy_change = energy_end - energy_baseline)

  county_change
}

county_energy <- load_egrid_year(analysis_year, egrid_change_baseline_year, energy_var)
```

5. Build Analysis Datasets
We use `county_energy` (which already contains `energy_change`) as the base dataset, then join income data.
```{r}
analysis_county_energy <- county_energy |>
  inner_join(
    income_aligned_app |>
      select(GEOID, median_income) |>
      rename(median_income_aligned = median_income),
    by = c("fips" = "GEOID")
  ) |>
  left_join(
    income_lagged_app |>
      select(GEOID, median_income) |>
      rename(median_income_lagged = median_income),
    by = c("fips" = "GEOID")
  )

```

6. Change in energy variable vs median household income
We take the selected eGRID energy variable at `egrid_change_baseline_year` and at `analysis_year`, compute county-level change inside `load_egrid_year()`, and work with the resulting `energy_change` column.
```{r}
analysis_energy_change <- analysis_county_energy |>
  drop_na(median_income_aligned, energy_change)
```

```{r}
cat("\n============================\n")
cat(
  "CHANGE ANALYSIS: selected energy variable (",
  egrid_change_baseline_year,
  " → ",
  analysis_year,
  ") vs median household income (ACS ",
  analysis_year,
  ", B19013_001)\n",
  sep = ""
)
cat("============================\n")

cor_change <- cor.test(
  analysis_energy_change$median_income_aligned,
  analysis_energy_change$energy_change,
  method = "pearson"
)
print(cor_change)

model_change <- lm(energy_change ~ median_income_aligned, data = analysis_energy_change)
print(summary(model_change))
```

```{r}
ggplot(analysis_energy_change, aes(x = median_income_aligned, y = energy_change)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = paste0(
      "Change in selected energy variable (",
      egrid_change_baseline_year,
      "–",
      analysis_year,
      ") vs median household income (",
      analysis_year,
      " ACS)"
    ),
    x = "Median household income (B19013_001)",
    y = energy_label
  ) +
  theme_minimal()
```

7. Choropleth: change in selected energy variable
Fill is `energy_change` from `analysis_energy_change`. A light basemap of all counties in the selected states is drawn underneath for surrounding context.

```{r map-nf-share-change-build}
state_fips <- geography_df |>
  mutate(state_fips = str_sub(fips, 1, 2)) |>
  distinct(state_fips) |>
  pull(state_fips)

counties_sf <- counties(state = state_fips, cb = TRUE, year = 2020) |>
  st_transform(crs = 5070)

counties_map_change <- counties_sf |>
  inner_join(
    analysis_energy_change |>
      select(fips, energy_change),
    by = c("GEOID" = "fips")
  )
```

```{r map-nf-share-change-plot}
ggplot() +
  geom_sf(data = counties_sf, fill = "grey95", color = "grey80", linewidth = 0.05) +
  geom_sf(data = counties_map_change, aes(fill = energy_change), color = "grey35", linewidth = 0.08) +
  scale_fill_gradient2(
    low = "#3B8132",
    mid = "#f7f7f7",
    high = "#7F2B0A",
    midpoint = 0,
    name = "Change in\nselected energy variable",
    labels = scales::label_comma()
  ) +
  theme_void() +
  theme(legend.position = "right") +
  labs(
    title = paste0(
      "Change in selected energy variable, ",
      egrid_change_baseline_year,
      "–",
      analysis_year
    ),
    subtitle = "Colored counties are analysis counties; grey counties provide surrounding-area context",
    caption = "Fill uses analysis_energy_change$energy_change."
  )
```


8. Partisanship vs change in selected energy variable

```{r partisanship-join-and-plot}
# Load .rda into an isolated environment so we can inspect objects safely
partisan_env <- new.env(parent = emptyenv())
load("NaNDA partisanship.rda", envir = partisan_env)

partisan_candidates <- ls(partisan_env)
partisan_tbl <- NULL

for (nm in partisan_candidates) {
  obj <- get(nm, envir = partisan_env)
  if (is.data.frame(obj) && "PARTISAN_INDEX_DEM" %in% names(obj)) {
    partisan_tbl <- obj
    break
  }
}

if (is.null(partisan_tbl)) {
  stop("No data frame with PARTISAN_INDEX_DEM found in NaNDA partisanship.rda")
}

fips_col_candidates <- c("STCOFIPS10", "FIPS", "GEOID", "geoid", "county_fips", "COUNTY_FIPS")
fips_col <- intersect(fips_col_candidates, names(partisan_tbl))

if (length(fips_col) == 0) {
  stop("Could not find a county ID column (e.g., fips/GEOID) in partisanship data")
}

partisan_join <- partisan_tbl |>
  transmute(
    fips = str_pad(as.character(.data[[fips_col[1]]]), 5, pad = "0"),
    PARTISAN_INDEX_DEM = as.numeric(PARTISAN_INDEX_DEM)
  )

analysis_partisanship <- analysis_energy_change |>
  left_join(partisan_join, by = "fips")

ggplot(
  analysis_partisanship |> drop_na(PARTISAN_INDEX_DEM, energy_change),
  aes(x = PARTISAN_INDEX_DEM, y = energy_change)
) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    labels = scales::label_number(accuracy = 0.1)
  ) +
  labs(
    title = paste0(
      "Partisan index vs change in selected energy variable (",
      egrid_change_baseline_year,
      "–",
      analysis_year,
      ")"
    ),
    x = "PARTISAN_INDEX_DEM (0 to 1)",
    y = energy_label
  ) +
  theme_minimal()
```

9. Metro vs micro vs rural (Census CBSA delineations)

```{r metro-micro-mean-sd-plot}
# Get metro/micro geographies from ACS via tidycensus
cbsa_sf <- get_acs(
  geography = "metropolitan statistical area/micropolitan statistical area",
  variables = "B01001_001",
  year = analysis_year,
  survey = "acs5",
  geometry = TRUE
) |>
  st_transform(crs = 5070) |>
  mutate(
    cbsa_type = case_when(
      str_detect(NAME, "Metro Area") ~ "Metropolitan",
      str_detect(NAME, "Micro Area") ~ "Micropolitan",
      TRUE ~ NA_character_
    )
  ) |>
  select(cbsa_type)

# Classify counties by whether their interior point lies in a metro/micro CBSA polygon
county_cbsa <- counties_sf |>
  st_point_on_surface() |>
  st_join(cbsa_sf, join = st_within, left = TRUE) |>
  st_drop_geometry() |>
  transmute(
    fips = GEOID,
    cbsa_type = replace_na(cbsa_type, "Rural")
  )

analysis_cbsa <- analysis_energy_change |>
  left_join(county_cbsa, by = "fips") |>
  mutate(
    county_type = cbsa_type,
    county_type = factor(county_type, levels = c("Metropolitan", "Micropolitan", "Rural"))
  )

cbsa_summary <- analysis_cbsa |>
  group_by(county_type) |>
  summarise(
    n = n(),
    mean_change = mean(energy_change, na.rm = TRUE),
    sd_change = sd(energy_change, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(cbsa_summary, aes(x = county_type, y = mean_change, fill = county_type)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_errorbar(
    aes(ymin = mean_change - sd_change, ymax = mean_change + sd_change),
    width = 0.18,
    linewidth = 0.6
  ) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_manual(values = c("Metropolitan" = "#E63400", "Micropolitan" = "#FF8B0F", "Rural" = "#6C3400")) +
  labs(
    title = paste0(
      "Mean change in selected energy variable by county CBSA type, ",
      egrid_change_baseline_year,
      "–",
      analysis_year
    ),
    subtitle = "Error bars show +/- 1 standard deviation",
    x = NULL,
    y = energy_label,
    caption = "CBSA classification from ACS metro/micro geographies via tidycensus."
  ) +
  theme_minimal()
```
